Online Social Cohesion Reflects Real-World Group Action in Syria during the Arab Spring

### 2020-05-29 manuscript.Rmd
Impose hard wrapping to improve versioning informativeness

abstract: |
  Social mobilization requires effective collective coordination. In recent years, individuals involved in social movements have taken to social media platforms to reach broader audiences at rapid speeds. Effective virtual mobilization similarly requires social cohesion and coordination, yet the degree to which virtual mobilization reflects or drives real-world movements is unclear. Here, we explore the dynamics of real-world events and social cohesion on Twitter in Syria during the Arab Spring. Using the nonlinear methods cross-recurrence quantification analysis and windowed cross-recurrence quantification analysis, we investigate if counts of all daily events, daily positive events, and daily negative events are coupled with social cohesion found in Syrian tweets. Results indicate that virtual social cohesion is coupled with the counts of all and negative events each day but not with counts of positive events. This suggests that virtual mobilization does reflect real-world movements but only for salient, frequent, and negative events. We conclude with a discussion of implications and applications of nonlinear methods in political science research.
  <!-- https://tinyurl.com/ybremelq -->

@@

## Use of Nonlinear Methods in Naturally Occurring Data
Traditional linear analyses have led to many notable discoveries in the domains of psychology and political science, but continuous data pose problems that are not easily overcome with these methods. Events have historically been conceptualized as linear chains of smaller, sequential events with one occurrence leading to the next. This notion has been increasingly replaced by ideas from dynamical systems theory. Rather than a chain of events, continuous data can be thought of as nested occurrences, where mutually constrained parts (e.g., citizens, governments, real-world events, social media platforms) are constantly interacting, leading to more global behaviors that are larger than the sum of the parts [@turvey2018lectures]. When analyzed as dynamical systems, many phenomena exhibit new nonlinear patterns that were formerly overlooked. Continuous data subject to linear analyses becomes flattened, and the variability that is fundamental to these systems and their evolution over time is ignored. Nonlinear methods provide means for capturing the rich temporal dynamics and variability that unfold in continuous data.
The primary methods used in the current work is cross-recurrence quantification analysis [CRQA; @zbilut1998detecting] and it’s off-shoot, windowed cross-recurrence quantification analysis [WCRQA; @webber2005recurrence]. CRQA has successfully been used to gain a better understanding of many complex human phenomena, such as human social interactions [for a review, see @fusaroli2014analyzing] and interpersonal synchrony [@paxton2017interpersonal]. We argue that the use of nonlinear methods in the investigation of political science phenomena will lead to new insights and development of more robust theories.

@@

The corpus of tweets used included `r sum(cohesion_df$n)` Syrian tweets occurring between March 30, 2012 and June 15, 2012. Using the Twitter streaming API, servers at the University of Texas at Austin collected the tweets. The servers were run on a stable (LTS) version of Ubuntu Linus, 10.04, and were connected directly to the high-bandwidth UT network maintained by IT services.
Tweets collected contained one or more of the following keywords: “syria,” “syrian,” “damascus,” “homs,” “al-assad,” and “sunni.” These keywords were chosen based on the most distinctive terms used in news articles covering the Arab Spring in Syria at the time. A node.js script was used to reinitialize the connection to Twitter's servers whenever the script timed out due to a period of silence. The script requested data from the API endpoint using basic HTTP authentication (OAuth was optional, at the time) at https://stream.twitter.com/1.1/statuses/filter.json, and stored the data directly to the local hard disk. The tweets were then post-processed with a Python script to convert the JSON-formatted tweets to tab-separated files and were then compressed to maximize storage efficiency.
All scripts used for scrapping tweets and compressing data can be found at https://github.com/chbrown/twilight.

@@

## Data Preparation
### Social Cohesion Metric
The social cohesion or linguistic similarity of daily tweets was calculated by quantifying the frequency of shared content words with respect to those words’ frequencies in the entire tweet corpus. If a linguistic term has a high frequency in one tweet but has a low frequency in the whole tweet corpus, then it is more likely a content word and is this given more weight in calculating social cohesion.
The tweets were first sorted into ascending order by timestamp. They were then grouped into successive sets of five tweets. Inverse frequency weighting for the overlap words and all words was conducted for each pair of two tweets in the group of five, then divided to get the social cohesion score for that pair (see Eq. 1-3). This resulted in ten social cohesion values for each group of five tweets, which were then averaged. Finally, the mean of all average cohesion values for all groups was taken to get a social cohesion score for each day.

@@

### ICEWS Event Time Series
Using the Integrated Crisis Early Warning System (ICEWS) developed at the Defense Advanced Research Project Agency, the daily events in Syria were scored with an intensity rating ranging from -10 to 10. Negative events, such as riots and attacks, are coded with low negative numbers while positive, peaceful events are coded with high positive numbers. This well-validated coding scheme allows for numerical quantification of different event types.
For both the target only and the target and source data, time series of the count of all events were generated by summing the total number of events on each day, irrespective of ICEWS score. Similarly, time series of count of positive events were generated by summing the number of ICEWS events for each day that had an intensity rating greater than zero. Lastly, time series of count of negative events were generated by summing the number of ICEWS events for each day that had an intensity rating less than zero.

## Analyses
### Cross-Recurrence Quantification Analysis
We first used cross-recurrence quantification analysis [CRQA; @zbilut1998detecting] to investigate the temporal patterns of social cohesion and real-world events in Syria during Arab Spring. As an extension of RQA, which captures the patterns of revisited states of a single time series [@zbilut1992embeddings; @webber1994dynamical], CRQA allows for two time series to be projected onto one another to investigate shared states and trajectories. Nonlinear analyses such as CRQA can reveal dynamics that are not captured by traditional linear methods. In order to perform CRQA, the time series need to be converted to the same scale, while still preserving a sampling rate that respects the native timescale of the system [cf. Nyquist frequency in signal processing; @grenander1959probability]. We converted each time series (e.g., Figure \@ref(fig:raw-ts)) into deciles (e.g., Figure \@ref(fig:deciled-ts)) to analyze the dynamics of different levels of social cohesion and counts of all, positive, and negative events on each day. In CRQA, a recurrent point is when two signals share the same state. These shared states can be at the same time or at later points in the time series.

@@

#### Metrics
The percentage of recurrent points to total possible points is a metric called recurrence rate (RR). This metrics gives a general description of how frequently the time series are sharing the same states. In the current study, a recurrent point would indicate that the level of social cohesion and frequency of events were at the same intensity.
Recurrent points that occur in successive moments form line structures. These are conceptualized as shared trajectories, where the two time series are moving together across time in a shared state. The percentage of recurrent points that make up diagonal line structures (i.e., two or more consecutive points) is a metric called percent determinism (DET). For the current project, a high DET would indicate that not only are the levels of social cohesion and frequency of events occurring at similar intensities, but they are doing so across consecutive days.
The average length of the line structures is called mean line (L). This can be conceptualized as the average amount of time that the two time series move together before one or both shift states. The longest line length is called max line (maxL). maxL indicates the longest instance of being “stuck” in a particular state together. The total number of lines is called NRLINE. The more lines there are, the more coordinated the system may be. The complexity or variability of line lengths is captured by a metric called entropy (ENTR) and entropy while accounting for the total number of lines in the plot is normalized entropy (rENTR). These entropy metrics help indicate the stability and predictability of the system. When one of the signals gets “stuck” in a state and the other moves on to other states, this results in vertical line structures. The proportion of points forming vertical line structures is a metric called laminarity (LAM) and the average length of the vertical trajectories is called trapping time (TT). For more detail on CRQA metrics, see @R-crqa.

@@

## Windowed-Recurrence Quantification Analysis
### Syria as Target
The time period of differing dynamics, from observation 38 through 43, appears to be captured by WCRQA of social cohesion and count of negative events for both RR and DET (Figure \@ref(fig:plot-wcrqa-targ-all)). The amount of recurrence and determinism for these periods almost consistently lie above chance, suggesting that the intensity of negative events coincide with the degree of social cohesion found on Twitter, and that these states carry across multiple days in a way that is different from that of chance.
The above-chance levels of determinism and recurrence rate are demonstrating the coupling of social cohesion and counts of all and negative events during times of strife. We can see this string of highly violent events reflected in the mode ICEWS event for each day (Figure \@ref(fig:plot-mode-event)), where from May 1st, 2020 to May 15th, 2020, the most frequently occurring event was almost consistently -10, the most violent event in the coding system. These dates directly correlate with the above chance RR and DET metrics for the WCRQA of social cohesion and count of negative events. Furthermore, for WCRQA of social cohesion and positive event count during this time period, DET was entirely within chance (Figure \@ref(fig:plot-wcrqa-targ-all)). This supports our hypothesis, as the positive events are seemingly not coupled with the social cohesion on Twitter during civil uprisings during the Arab Spring.

>>>>

abstract: |
  Social mobilization requires effective collective coordination. In recent years, individuals involved in social movements have taken to social media platforms to reach broader audiences at rapid speeds. Effective virtual mobilization similarly requires social cohesion and coordination, yet the degree to which virtual mobilization reflects or drives real-world movements is unclear. Here, we explore the dynamics of real-world events and social cohesion on Twitter in Syria during the Arab Spring. Using the nonlinear methods cross-recurrence quantification analysis and windowed cross-recurrence quantification analysis, we investigate if counts of all daily events, daily positive events, and daily negative events are coupled with social cohesion found in Syrian tweets. Results indicate that virtual social cohesion is coupled with the counts of all and negative events each day but not with counts of positive events. This suggests that virtual mobilization does reflect real-world movements but only for salient, frequent and negative events. We conclude with a discussion of implications and applications of nonlinear methods in political science research.
  <!-- https://tinyurl.com/ybremelq -->

@@

## Use of Nonlinear Methods in Naturally Occurring Data
Traditional linear analyses have led to many notable discoveries in the 
domains of psychology and political science, but continuous data pose 
problems that are not easily overcome with these methods. Events have
historically been conceptualized as linear chains of smaller, sequential 
events with one occurrence leading to the next. This notion has been 
increasingly replaced by ideas from dynamical systems theory. Rather than 
a chain of events, continuous data can be thought of as nested occurrences, 
where mutually constrained parts (e.g., citizens, governments, real-world 
events, social media platforms) are constantly interacting, leading to 
more global behaviors that are larger than the sum of the parts 
[@turvey2018lectures]. When analyzed as dynamical systems, many phenomena 
exhibit new nonlinear patterns that were formerly overlooked. Continuous 
data subject to linear analyses becomes flattened, and the variability 
that is fundamental to these systems and their evolution over time is 
ignored. Nonlinear methods provide means for capturing the rich temporal 
dynamics and variability that unfold in continuous data.

The primary methods used in the current work is cross-recurrence 
quantification analysis [CRQA; @zbilut1998detecting] and its offshoot, 
windowed cross-recurrence quantification analysis [WCRQA; @webber2005recurrence]. 
CRQA has successfully been used to gain a better understanding of many 
complex human phenomena, such as human social interactions 
[for a review, see @fusaroli2014analyzing] and interpersonal synchrony 
[@paxton2017interpersonal]. We argue that the use of nonlinear methods in 
the investigation of political science phenomena will lead to new insights 
and development of more robust theories.

@@

The corpus of tweets used included `r sum(cohesion_df$n)` Syrian tweets 
occurring between March 30, 2012 and June 15, 2012. Using the Twitter 
streaming API, servers at the University of Texas at Austin collected 
the tweets. The servers were run on a stable (LTS) version of Ubuntu 
Linus, 10.04, and were connected directly to the high-bandwidth UT network 
maintained by IT services.

Tweets collected contained one or more of the following keywords:
“syria,” “syrian,” “damascus,” “homs,” “al-assad,” and “sunni.” These 
keywords were chosen based on the most distinctive terms used in news 
articles covering the Arab Spring in Syria at the time. A node.js script 
was used to reinitialize the connection to Twitter's servers whenever the 
script timed out due to a period of silence. The script requested data from 
the API endpoint using basic HTTP authentication (OAuth was optional, at 
the time) at https://stream.twitter.com/1.1/statuses/filter.json, and stored 
the data directly to the local hard disk. The tweets were then post-processed
with a Python script to convert the JSON-formatted tweets to tab-separated 
files and were then compressed to maximize storage efficiency.

All scripts used for scrapping tweets and compressing data can be found at https://github.com/chbrown/twilight.

@@

## Data Preparation
### Social Cohesion Metric
The social cohesion or linguistic similarity of daily tweets was calculated
by quantifying the frequency of shared content words with respect to those 
words’ frequencies in the entire tweet corpus. If a linguistic term has a 
high frequency in one tweet but has a low frequency in the whole tweet 
corpus, then it is more likely a content word and is this given more weight
in calculating social cohesion.

The tweets were first sorted into ascending order by timestamp. They were 
then grouped into successive sets of five tweets. Inverse frequency 
weighting for the overlap words and all words was conducted for each pair 
of two tweets in the group of five, then divided to get the social cohesion 
score for that pair (see Eq. 1-3). This resulted in ten social cohesion 
values for each group of five tweets, which were then averaged. Finally, 
the mean of all average cohesion values for all groups was taken to get 
a social cohesion score for each day.

@@

### ICEWS Event Time Series
Using the Integrated Crisis Early Warning System (ICEWS) developed at the 
Defense Advanced Research Project Agency, the daily events in Syria were 
scored with an intensity rating ranging from -10 to 10. Negative events, 
such as riots and attacks, are coded with low negative numbers while 
positive, peaceful events are coded with high positive numbers. This 
well-validated coding scheme allows for numerical quantification of 
different event types.

For both the target only and the target and source data, time series of 
the count of all events were generated by summing the total number of events 
on each day, irrespective of ICEWS score. Similarly, time series of count 
of positive events were generated by summing the number of ICEWS events 
for each day that had an intensity rating greater than zero. Lastly, 
time series of count of negative events were generated by summing the number 
of ICEWS events for each day that had an intensity rating less than zero.

## Analyses
### Cross-Recurrence Quantification Analysis
We first used cross-recurrence quantification analysis 
[CRQA; @zbilut1998detecting] to investigate the temporal patterns of 
social cohesion and real-world events in Syria during Arab Spring. As 
an extension of RQA, which captures the patterns of revisited states of 
a single time series [@zbilut1992embeddings; @webber1994dynamical], CRQA 
allows for two time series to be projected onto one another to investigate 
shared states and trajectories. Nonlinear analyses such as CRQA can reveal
dynamics that are not captured by traditional linear methods.

In order to perform CRQA, the time series need to be converted to the same 
scale, while still preserving a sampling rate that respects the native 
timescale of the system 
[cf. Nyquist frequency in signal processing; @grenander1959probability]. 
We converted each time series (e.g., Figure \@ref(fig:raw-ts)) into deciles
(e.g., Figure \@ref(fig:deciled-ts)) to analyze the dynamics of different 
levels of social cohesion and counts of all, positive, and negative events 
on each day. In CRQA, a recurrent point is when two signals share the same 
state. These shared states can be at the same time or at later points in 
the time series.

@@

#### Metrics
The percentage of recurrent points to total possible points is a metric 
called recurrence rate (RR). This metrics gives a general description of 
how frequently the time series are sharing the same states. In the current 
study, a recurrent point would indicate that the level of social cohesion 
and frequency of events were at the same intensity.

Recurrent points that occur in successive moments form line structures. 
These are conceptualized as shared trajectories, where the two time series 
are moving together across time in a shared state. The percentage of 
recurrent points that make up diagonal line structures (i.e., two or more 
consecutive points) is a metric called percent determinism (DET). For the 
current project, a high DET would indicate that not only are the levels of 
social cohesion and frequency of events occurring at similar intensities, 
but they are doing so across consecutive days.

The average length of the line structures is called mean line (L). This can 
be conceptualized as the average amount of time that the two time series move
together before one or both shift states. The longest line length is called 
max line (maxL). maxL indicates the longest instance of being “stuck” in a
particular state together. The total number of lines is called NRLINE. The 
more lines there are, the more coordinated the system may be. The 
complexity or variability of line lengths is captured by a metric called 
entropy (ENTR) and entropy while accounting for the total number of lines in 
the plot is normalized entropy (rENTR). These entropy metrics help indicate 
the stability and predictability of the system. When one of the signals gets
“stuck” in a state and the other moves on to other states, this results in
vertical line structures. The proportion of points forming vertical line 
structures is a metric called laminarity (LAM) and the average length of the
vertical trajectories is called trapping time (TT). For more detail on CRQA 
metrics, see @R-crqa.

@@

## Windowed-Recurrence Quantification Analysis
### Syria as Target
The time period of differing dynamics, from observation 38 through 43, 
appears to be captured by WCRQA of social cohesion and count of negative 
events for both RR and DET (Figure \@ref(fig:plot-wcrqa-targ-all)). The 
amount of recurrence and determinism for these periods almost consistently 
lie above chance, suggesting that the intensity of negative events coincide
with the degree of social cohesion found on Twitter, and that these states 
carry across multiple days in a way that is different from that of chance.

The above-chance levels of determinism and recurrence rate are demonstrating
the coupling of social cohesion and counts of all and negative events during
times of strife. We can see this string of highly violent events reflected 
in the mode ICEWS event for each day (Figure \@ref(fig:plot-mode-event)), 
where from May 1st, 2020 to May 15th, 2020, the most frequently occurring 
event was almost consistently -10, the most violent event in the coding 
system. These dates directly correlate with the above chance RR and DET 
metrics for the WCRQA of social cohesion and count of negative events. 
Furthermore, for WCRQA of social cohesion and positive event count during 
this time period, DET was entirely within chance (Figure 
\@ref(fig:plot-wcrqa-targ-all)). This supports our hypothesis, as the 
positive events are seemingly not coupled with the social cohesion on Twitter 
during civil uprisings during the Arab Spring.
